---
title: "AP-1 Cellular Memory: Kinase Analysis"
subtitle: "Reproducible analysis of persistent kinase expression and interference responses"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    self_contained: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

options(
  digits = 4,
  scipen = 4
)
```

# Overview

This report documents a reproducible computational analysis of kinase expression
associated with cellular memory in the AP-1 cellular adaptation model.

The analysis is based on the study:

**Li J, Ravindran PT, O'Farrell A, et al.
AP-1 mediates cellular adaptation and memory formation.
Nature Communications 17, 4265 (2026).**

The goal of this downstream analysis is to identify kinase genes associated with
the post-dexamethasone memory state and examine how those kinase responses behave
under AP-1-related interference conditions.

# Reproducible workflow

The complete workflow is executed with:

```text
Rscript run_all.R
```

The main analysis pipeline is:

```text
00_fetch_data.R
        ↓
01_setup_and_data.R
        ↓
02_initial_Dex_response.R
        ↓
03_persistent_kinase_memory.R
        ↓
04_kinase_interference.R
        ↓
05_integrated_kinase_trajectory.R
        ↓
06_make_final_kinase_figure.R
```

The workflow automatically retrieves the required public input files when the
`data/` directory is missing.

# Input data

The analysis uses:

```text
data/log2CPM_matrix.txt
data/studydesign.tsv
data/dex_log2FC.txt
data/dextable.txt
data/KinHubKinaseList.csv
```

The AP-1 RNA-seq files originate from the public cellular-memory repository:

<https://github.com/arjunrajlaboratory/cellularmemory>

Relevant source directory:

```text
extractedData/burninrnaseq/202208/
```

The kinase reference list is obtained from the KinHub/OpenKinome resource.

# Load analysis outputs

```{r load-results}
required_outputs <- c(
  "results/Initial_Dex_kinase_response.csv",
  "results/Memory_9_significant_kinases.csv",
  "results/Memory_9_kinases_interference.csv",
  "results/Integrated_9_kinase_trajectory.csv"
)

missing_outputs <- required_outputs[
  !file.exists(required_outputs)
]

if (length(missing_outputs) > 0) {
  stop(
    "Required result file(s) missing: ",
    paste(missing_outputs, collapse = ", "),
    ". Run Rscript run_all.R before rendering this report."
  )
}

initial <- read.csv(
  "results/Initial_Dex_kinase_response.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

memory <- read.csv(
  "results/Memory_9_significant_kinases.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

interference <- read.csv(
  "results/Memory_9_kinases_interference.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

integrated <- read.csv(
  "results/Integrated_9_kinase_trajectory.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE
)
```

# Data integrity and sample alignment

The expression dataset contains 191 analyzable samples after alignment with the
experimental design.

`Sample_147` is present in the design metadata but is absent from the final
expression matrix and is therefore excluded during the alignment step.

The setup script verifies that the final metadata order exactly matches the
expression-matrix sample order before downstream analysis.

# Initial dexamethasone kinase response

The initial dexamethasone analysis evaluates kinase-associated transcriptional
changes during the initial response phase.

```{r initial-summary}
cat("Kinases evaluated:", nrow(initial), "\n")

cat(
  "Kinases with kinase-level FDR < 0.05:",
  sum(initial$kinase_FDR < 0.05, na.rm = TRUE),
  "\n"
)
```

```{r initial-table}
initial_show <- initial[
  order(initial$kinase_FDR),
  c(
    "kinase",
    "logFC",
    "P.Value",
    "adj.P.Val",
    "kinase_FDR"
  )
]

knitr::kable(
  head(initial_show, 15),
  digits = 4,
  caption = "Top initial dexamethasone-associated kinase responses"
)
```

# Persistent kinase memory

The post-dexamethasone memory analysis identified a set of kinase candidates
passing the kinase-level FDR threshold.

```{r persistent-list}
cat(
  paste(memory$kinase, collapse = ", ")
)
```

```{r memory-table}
memory_show <- memory[
  ,
  c(
    "kinase",
    "logFC",
    "P.Value",
    "adj.P.Val",
    "kinase_FDR"
  )
]

knitr::kable(
  memory_show,
  digits = 4,
  caption = "FDR-significant kinase candidates in the memory contrast"
)
```

# Integrated kinase trajectories

The integrated analysis compares each persistent kinase across:

- initial dexamethasone response;
- post-dexamethasone memory;
- memory with JNK-IN-8;
- memory with T-5224;
- JNK-IN-8 interaction;
- T-5224 interaction.

```{r integrated-table}
integrated_show <- integrated[
  ,
  c(
    "kinase",
    "Initial_Dex_logFC",
    "Initial_Dex_kinaseFDR",
    "Memory_logFC",
    "Memory_kinaseFDR",
    "JNKIN8_Memory_logFC",
    "T5224_Memory_logFC",
    "Trajectory"
  )
]

knitr::kable(
  integrated_show,
  digits = 4,
  caption = "Integrated trajectory of the nine memory-associated kinase candidates"
)
```

# Trajectory classification

```{r trajectory-summary}
trajectory_summary <- as.data.frame(
  table(integrated$Trajectory),
  stringsAsFactors = FALSE
)

colnames(trajectory_summary) <- c(
  "Trajectory",
  "Number_of_kinases"
)

knitr::kable(
  trajectory_summary,
  caption = "Kinase trajectory classes"
)
```

The trajectory classes distinguish kinase changes already detectable during the
initial dexamethasone response from changes that become more prominent in the
memory state.

# AP-1-related interference analysis

The interference analysis evaluates how the memory-associated kinase effects
behave in the presence of JNK-IN-8 and T-5224.

```{r interference-table}
interference_show <- interference[
  ,
  c(
    "kinase",
    "No_Inhib_logFC",
    "JNKIN8_logFC",
    "JNK_interaction",
    "JNK_interaction_P",
    "JNK_candidate_FDR",
    "T5224_logFC",
    "T5224_interaction",
    "T5224_interaction_P",
    "T5224_candidate_FDR"
  )
]

knitr::kable(
  interference_show,
  digits = 4,
  caption = "Kinase responses under AP-1-related interference conditions"
)
```

# Final kinase trajectory figure

```{r final-figure, out.width="100%"}
figure_file <- "figures/Final_Kinase_Memory_Interference.png"

if (!file.exists(figure_file)) {
  stop(
    "Final figure is missing. Run Rscript run_all.R first."
  )
}

knitr::include_graphics(
  figure_file
)
```

**Figure.** Expression effect-size profiles of the nine significant
memory-associated kinase candidates across the initial dexamethasone response,
memory state, JNK-IN-8 interference, and T-5224 interference conditions.

# Key output files

The main outputs generated by the pipeline are:

```text
results/Initial_Dex_kinase_response.csv
results/Memory_all_476_kinases.csv
results/Memory_9_significant_kinases.csv
results/Memory_9_kinases_interference.csv
results/Integrated_9_kinase_trajectory.csv

figures/Final_Kinase_Memory_Interference.png
figures/Final_Kinase_Memory_Interference.pdf
```

# Reproducibility

The repository uses `renv` to record the R package environment.

A fresh clone can be reproduced with:

```text
git clone <repository-url>
cd AP1_cellular_memory_analysis
```

Restore the package environment:

```text
Rscript -e "renv::restore()"
```

Then execute the complete workflow:

```text
Rscript run_all.R
```

The data-fetching script automatically downloads the required external data
when they are not already present.

# Software environment

```{r software-environment}
cat(
  "R version:",
  R.version.string,
  "\n"
)

cat(
  "Platform:",
  R.version$platform,
  "\n"
)

packages <- c(
  "limma",
  "dplyr",
  "ggplot2",
  "rmarkdown",
  "knitr"
)

package_versions <- data.frame(
  Package = packages,
  Version = vapply(
    packages,
    function(x) {
      if (requireNamespace(x, quietly = TRUE)) {
        as.character(
          packageVersion(x)
        )
      } else {
        "Not installed"
      }
    },
    character(1)
  )
)

knitr::kable(
  package_versions,
  caption = "Software packages used in the analysis and report"
)
```

# Summary

This reproducible workflow:

1. retrieves the required public datasets;
2. validates sample alignment;
3. quantifies the initial dexamethasone kinase response;
4. identifies persistent kinase-associated memory;
5. evaluates JNK-IN-8 and T-5224 interference conditions;
6. integrates kinase trajectories;
7. generates the final kinase trajectory figure.

The current analysis identifies nine kinase candidates that pass the defined
memory-associated kinase FDR criterion.

Their initial-response, memory, and interference profiles are retained in the
integrated result table for downstream biological interpretation.

# ATAC-seq integration of the nine kinase candidates

An additional chromatin-accessibility analysis was performed for the nine
FDR-significant memory-associated kinase candidates using the processed ATAC-seq
data from the original AP-1 cellular-memory study.

Gene-body plus 2 kb upstream bigWig accessibility was summarized across:

- 3-day DMSO;
- 3-day Dex;
- Dex followed by DMSO;
- DMSO followed by trametinib;
- Dex followed by trametinib.

These regional bigWig summaries are used as descriptive accessibility
trajectories and are not treated as formal differential-accessibility tests.

For formal peak-level evidence, the authors' DESeq2 differential ATAC results
for Dex-to-Tram versus DMSO-to-Tram were intersected with the nine kinase
regions.

```{r atac-peak-statistics}
atac_stats_file <- "results/ATAC_integration/Final_9_kinase_peak_statistics.tsv"

if (!file.exists(atac_stats_file)) {
  stop("Missing ATAC result file: ", atac_stats_file)
}

atac_stats <- read.delim(
  atac_stats_file,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

atac_show <- atac_stats[
  ,
  c(
    "kinase",
    "RNA_memory_logFC",
    "RNA_memory_FDR",
    "memory_peak_count",
    "memory_significant_peak_count",
    "best_memory_peak",
    "peak_log2FC",
    "peak_FDR",
    "final_ATAC_interpretation"
  )
]

knitr::kable(
  atac_show,
  digits = 4,
  caption = "Peak-level ATAC evidence for the nine memory-associated kinase candidates"
)
```

Across the nine kinase candidates, six reported differential ATAC peaks
overlapped five kinase loci. Only one overlapping peak passed FDR < 0.05.

The significant peak was associated with **CDK2**:

```text
chr12:56358987-56360175
DESeq2 log2FoldChange = +1.028
adjusted P value = 0.00349
```

CDK2 RNA expression in the memory contrast was significantly decreased
(logFC approximately -0.881; FDR approximately 0.0226), whereas accessibility
at this local ATAC peak increased. Thus, the RNA and chromatin-accessibility
responses are discordant.

The CDK2 peak was not represented among the peaks reported in the authors'
3-day Dex versus DMSO differential ATAC table. It is therefore interpreted here
as a memory-associated / Dex-history-associated accessibility change rather
than as a canonical acute-Dex-induced learned peak.

## CDK2 significant ATAC peak trajectory

```{r cdk2-atac-figure, out.width="90%"}
cdk2_figure <- "figures/CDK2_significant_ATAC_peak_trajectory.png"

if (!file.exists(cdk2_figure)) {
  stop("Missing CDK2 ATAC figure: ", cdk2_figure)
}

knitr::include_graphics(cdk2_figure)
```

**Figure.** Accessibility trajectory at the statistically significant CDK2
memory-associated ATAC peak. Open circles represent biological replicates and
filled points represent condition means. The DESeq2 statistic refers specifically
to the Dex-to-Tram versus DMSO-to-Tram comparison.
